```{r setup, echo=FALSE, cache=FALSE}
options(scipen = 6, digits = 3)
```

# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

Load the activity data into a data.table (we like data.tables as they're fast), and convert the date column from characters to Dates.
```{r}
library(data.table)
actdata <- fread("activity/activity.csv")
actdata[,date:= as.Date( date, format = "%Y-%m-%d")]
```

## What is mean total number of steps taken per day?
Create a data.table with the sum of steps for each day, excluding any observations missing step data.
```{r, echo=TRUE}
dailySteps <- actdata[!is.na(steps),sum(steps), by=date]
setnames(dailySteps,"V1","daily.steps")
```
```{r, echo=FALSE}
## knitr won't preview a data.table based on a setnames() call, so we do it ourselves
print(dailySteps, topn=5)
```
### *Make a histogram of the total number of steps taken each day*
```{r dailystephist, echo=TRUE, tidy=FALSE}
hist(dailySteps$daily.steps, 
      breaks = 8,
      ylim = c(0,25),
      col = "cornflowerblue",
      main = "Histogram of Steps Per Day",
      xlab = "Steps Per Day",
      ylab = "Frequency (Days)")
```
### *Calculate and report the mean and median total number of steps taken per day*
```{r dailystepmeanmed, echo=TRUE}
meansteps <- mean(dailySteps$daily.steps)
meansteps
medsteps <- median(dailySteps$daily.steps)
medsteps
```
Mean steps per day: **`r meansteps`**

Median steps per day: **`r medsteps`**

## What is the average daily activity pattern?
Create a data.table with the mean of steps for each interval, excluding any observations missing step data.
```{r, echo=TRUE}
intervalSteps <- actdata[!is.na(steps),mean(steps), by=interval]
setnames(intervalSteps,"V1","mean.steps")
```
```{r, echo=FALSE}
## knitr won't preview a data.table based on a setnames() call, so we do it ourselves
print(intervalSteps, topn=5)
```

### *A time series plot of the 5-minute interval and the mean steps taken in that interval across all days.*
```{r meanstepsbyinterval, echo=TRUE, tidy=FALSE}
plot(x = intervalSteps$interval, y = intervalSteps$mean.steps,
      ylim = c(0, 300),
      type = "l",
      col = "cornflowerblue",
      main = "Mean Steps By Interval",
      xlab = "Interval",
      ylab = "Mean Steps")
```

### *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*
```{r, echo=TRUE}
maxinv <- intervalSteps[mean.steps == max(intervalSteps$mean.steps),interval]
maxinv
```
Interval **`r maxinv`** is has the largest average number of steps.

## Imputing missing values
### *Calculate and report the total number of missing values in the dataset*
```{r, echo=TRUE}
nrowwithna <- sum(is.na(actdata$steps))
nrowwithna
```
There are **`r nrowwithna`** rows with NA values.

### *Devise a strategy for filling in all of the missing values in the dataset.*
We fill in the missing steps values with the mean number of steps for the corresponding interval.

### *Create a new dataset that is equal to the original dataset but with the missing data filled in.*
A few notes on what this code is doing:

* The "by = interval" term creates a temporary sub-data.table for each unique interval.
* The ifelse() expression in the j term is run within the context of each sub-table. 
* .SD is a special reference to the sub-table, allowing us to determine the mean value of the steps column within each sub table. We set the value of any NA observations to that mean.

```{r, echo=TRUE, tidy=FALSE}
actImp <- copy(actdata)
actImp[,steps := ifelse( is.na(steps), 
                        as.integer(mean(.SD$steps, na.rm = TRUE)), 
                        steps),
        by = interval]
```

### *Histogram of the total number of steps taken each day*
Using the table with imputed values, create a data.table with the sum of steps for each day, then generate a histogram using that summary table.

```{r, echo=TRUE}
dailyImputed <- actImp[,sum(steps), by=date]
setnames(dailyImputed,"V1","daily.steps")
```
```{r, echo=FALSE}
## knitr won't preview a data.table based on a setnames() call, so we do it ourselves
print(dailyImputed, topn=5)
```
```{r dailyimputedhist, echo=TRUE, tidy=FALSE}
hist(dailyImputed$daily.steps, 
      breaks = 8,
      ylim = c(0,25),
      col = "cornflowerblue",
      main = "Histogram of Steps Per Day",
      xlab = "Steps Per Day",
      ylab = "Frequency (Days)")
```

### *Calculate and report the mean and median total number of steps taken per day.*
```{r, echo=TRUE}
meanimp <- mean(dailyImputed$daily.steps)
meanimp
medimp <- median(dailyImputed$daily.steps)
medimp
```
Mean steps per day (with imputed data): **`r meanimp`**

Median steps per day (with imputed data): **`r medimp`**

Mean steps per day: **`r meansteps`**

Median steps per day: **`r medsteps`**
### *Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?*
Imputing missing step count values does cause the mean and median daily step counts to differ, though not greatly. 

| NA Values   | Mean           | Median        |
| ----------- | -------------: | ------------: |
| Removed     | `r meansteps`  | `r medsteps`  |
| Imputed     | `r meanimp`    | `r medimp`    |


## Differences in activity patterns between weekdays and weekends
### *Create a new factor variable in the dataset indicating whether a given date is a weekday or weekend day.*
```{r, topn=8, echo=TRUE, tidy=FALSE}
actImp[, day.type := as.factor(
                        ifelse( (tolower(weekdays(date)) == "saturday") | 
                                (tolower(weekdays(date)) == "sunday"),
                              "weekend", 
                              "weekday"))]
levels(actImp$day.type)
```
We can check that the day assignment was correct with the following summary. topn=10 is used to ensure we show 5 weekdays bookended with 2 weekends.
```{r }
print(actImp[,unique(day.type), by = date], topn=10)
```
### *Plots of the 5-minute interval and the mean steps taken in that interval across weekend and weekdays.*
Let's create a summary table with the mean steps for each interval and day.type.
```{r, echo=TRUE}
stepsByType = actImp[, mean(.SD$steps), by=c("interval","day.type")]
setnames(stepsByType,"V1","mean.steps")
```
```{r, echo=FALSE}
## knitr won't preview a data.table based on a setnames() call, so we do it ourselves
print(stepsByType, topn=5)
```
Plot the summary table, enforcing a common y-axis scale to make comparisons between the plots easier.
```{r weekendweekdaysteps, echo=TRUE, tidy=FALSE}
par(mfrow = c(2,1))
plot(x = stepsByType[day.type == "weekend",interval],
      y = stepsByType[day.type == "weekend",mean.steps],
      ylim = c(0, 300),
      type = "l",
      col = "cornflowerblue",
      main = "Mean Steps By Interval - Weekend",
      xlab = "Interval",
      ylab = "Number of Steps")
plot(x = stepsByType[day.type == "weekday",interval],
      y = stepsByType[day.type == "weekday",mean.steps],
      ylim = c(0, 300),
      type = "l",
      col = "cornflowerblue",
      main = "Mean Steps By Interval - Weekday",
      xlab = "Interval",
      ylab = "Number of Steps")
```